The role of cross-linguistic lexical similarity on bilingual word acquisition

### 2022-09-13 manuscript/manuscript.Rmd
Update report

bibliography      : ["trajectories.bib"]

@@

knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE,
  cache.extra = knitr::rand_seed,
  out.width = "100%",
  dpi = 300,
  results = "asis"
)
```

@@

# load packages
library(papaja)
library(knitr)
library(here)
library(dplyr)
library(tidyr)
library(tidybayes)
library(scales)
library(janitor)
library(stringr)
library(ggplot2)
library(patchwork)

tar_load(multilex_data)
tar_load(items)
tar_load(participants)
tar_load(responses)
tar_load(loo_comp)
tar_load(loo_prod)
tar_load(fits_comp)
tar_load(fits_prod)

options(
  knitr.kable.NA = "-",
  knitr.duplicate.label = "allow",
  ggplot2.discrete.fill = wesanderson::wes_palettes$Darjeeling1,
  ggplot2.discrete.colour = wesanderson::wes_palettes$Darjeeling1,
  ggplot2.continuous.fill = ggplot2::scale_color_gradient,
  ggplot2.continuous.colour = ggplot2::scale_color_gradient
)

theme_set(theme_apa())

@@

```{r participants, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

n_participants <- length(unique(participants$id))
dates <- as_date(range(participants$time_stamp, na.rm = TRUE))

@@

```{r participants_lp, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Participant sample size by age and degree of exposure to a second language. Information about DoE was provided by families when filling the questionnaire. For illustration purposes this table, DoEs were binned into 10\\% wide bins, and ages (in months) were round to the nearest integer. Hyphens indicate that no participants from that specific combination of age filled the questionnaire."}
labs <- c("0-10%", "10-20%", "20-30%", "30-40%", "40-50%")
t <- participants %>% 
  mutate(
    age = round(age),
    doe_2 = ifelse(dominant_language=="Catalan", doe_spanish, doe_catalan),
    doe_2 = cut(doe_2, breaks = seq(0, 0.5, 0.1), include.lowest = TRUE, labels = labs)
  ) %>% 
  count(age, doe_2) %>% 
  arrange(doe_2) %>% 
  mutate(doe_2 = fct_inorder(doe_2)) %>% 
  pivot_wider(names_from = age, values_from = n, values_fill = NA) %>% 
  gt() %>% 
  tab_spanner(label = "Number of participants by age (months)", columns = 2:17) %>% 
  fmt_missing(everything(), everything(), missing_text = "-") %>% 
  cols_label(
    doe_2 = md("**L2 DoE**")
  ) %>% 
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels(everything())
  ) %>% 
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners(everything())
  )

gtsave(t, here("Figures/participants_lp.png"))

We use the highest educational attainment of parents or caretakers as a proxy of participants' socio-economic status (SES), which families self-reported in the questionnaire by filling two items asking for the educational attainment of each parent or caretaker, with the following available options: *No education*, *Primary*, *Secondary*, *Complementary*, *Vocational*, and *University*, in line with the current educational system in Spain. Most families reported university studies (`r table(participants$edu_parent)["University"]`, `r table(participants$edu_parent)["University"]/length(unique(participants$id))`), followed by families were the highest educational attainment were vocational studies (`r table(participants$edu_parent)["Vocational"]`, `r percent(table(participants$edu_parent)["Vocational"]/length(unique(participants$id)))`), complementary studies (`r table(participants$edu_parent)["Complementary"]`,  `r percent(table(participants$edu_parent)["Complementary"]/length(unique(participants$id)))`), secondary education (`r table(participants$edu_parent)["Secondary"]`, `r percent(table(participants$edu_parent)["Secondary"]/length(unique(participants$id)))`, `r percent(table(participants$edu_parent)["Primary"]/length(unique(participants$id)))`), no formal education (`r table(participants$edu_parent)["No education"]`, `r percent(table(participants$edu_parent)["No education"]/length(unique(participants$id)))` and primary education (`r table(participants$edu_parent)["Primary"]`, <1%). Highest educational attainment was similar across participants with different DoE to a second language, and therefore, differences in comprehension or production between participants with different language profiles are unlikely to stem from a difference in SES [@fernald2013ses] (see Table 2).

```{r participants_edu, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Educational attainment. Proportion of highest parental educational attainment by DoE to the second language. For illustration purposes, DoEs were binned into 10\\% wide bins."}
labs <- c("0-10%", "10-20%", "20-30%", "30-40%", "40-50%")

t <- participants %>%
  mutate(
    age = round(age),
    doe_2 = ifelse(dominant_language=="Catalan", doe_spanish, doe_catalan),
    doe_2 = cut(doe_2, breaks = seq(0, 0.5, 0.1), include.lowest = TRUE, labels = labs)
  ) %>% 
  group_by(doe_2) %>%
  summarise(
    n_no_education = length(edu_parent[edu_parent=="No education"])/n(),
    n_primary = length(edu_parent[edu_parent=="Primary"])/n(),
    n_secondary = length(edu_parent[edu_parent=="Secondary"])/n(),
    n_complementary = length(edu_parent[edu_parent=="Complementary"])/n(),
    n_vocational = length(edu_parent[edu_parent=="Vocational"])/n(),
    n_university = length(edu_parent[edu_parent=="University"])/n(),
    .groups = "drop"
  ) %>% 
  gt() %>% 
  tab_spanner("Educational attainment", columns = 2:7) %>% 
  cols_label(
    doe_2 = md("**L2 DoE**"),
    n_no_education = "No education",
    n_primary = "Primary",
    n_secondary = "Secondary",
    n_complementary = "Complementary",
    n_vocational = "Vocational",
    n_university = "University"
  ) %>% 
  fmt_percent(2:7) %>% 
  fmt_missing(everything(), missing_text = "-") %>% 
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels(everything())
  ) %>% 
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners(everything())
  )

gtsave(t, here("Figures/participants_edu.png"))

@@

```{r items, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
pool <- multilex::pool %>% 
  drop_na(cognate)
n_categories <- distinct(pool, category) %>% nrow()
n_items <- length(unique(pool$item))
n_item_language <- count(pool, language) %>% pull(n) %>% set_names(c("catalan", "spanish"))
n_item_cognate <- count(pool, language, cognate) %>% 
  pull(n) %>% 
  set_names(c("catalan-cognate", "catalan-noncognate", "spanish-cognate", "spanish-noncognate"))
```

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items (`r n_item_cognate["catalan-cognate"]` cognates, `r n_item_cognate["catalan-noncognate"]` non-cognates) and the Spanish inventory contained `r n_item_language["spanish"]` (`r n_item_cognate["spanish-cognate"]` cognates, `r n_item_cognate["spanish-noncognate"]` non-cognates). Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. When there were two acceptable translation equivalents for a given word, we included both in separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single items (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*. We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (800 translation equivalents; 800 items in Catalan, 800 items in Spanish), while others filled a shorter version (~400 translation equivalents, ~400 items in Catalan, ~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were carefully designed so that each contained a representative subsample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions. We excluded from the analysis multi-word items (e.g., *barrita de cereales* [cereal bar]) and items that included more than one word-form (e.g., *mono / mico*). Table 3 shows the classification of items in cognates and non-cognates and their lexical frequency scores across the four lists of the inventories.

```{r items_frequency, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Lexical frequencies. Mean, standard error, and 95\\% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."}

d <- items %>%
  select(te, language, cognate, frequency) %>% 
  left_join(select(multilex_data$pool, te, language, version)) %>% 
  unnest(version) %>% 
  group_by(language, version, cognate) %>% 
  summarise(
    frequency_mean = mean(frequency, na.rm = TRUE),
    frequency_sd = sd(frequency, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) %>% 
  mutate(
    frequency_se = frequency_sd/sqrt(n),
    frequency_ci_lower = frequency_mean + qnorm(0.025)*frequency_se,
    frequency_ci_upper = frequency_mean + qnorm(0.975)*frequency_se
  ) %>% 
  pivot_wider(names_from = c("cognate"), values_from = c("n", matches("frequency"))) %>% 
  clean_names() %>% 
  mutate(
    frequency_ci_identical_cognate = paste0(
      "[", round(frequency_ci_lower_identical_cognate, 2), ", ", round(frequency_ci_lower_identical_cognate, 2), "]"
    ),
    frequency_ci_non_identical_cognate = paste0(
      "[", round(frequency_ci_lower_non_identical_cognate, 2), ", ", round(frequency_ci_lower_non_identical_cognate, 2), "]"
    ),
    frequency_ci_non_cognate = paste0(
      "[", round(frequency_ci_lower_non_cognate, 2), ", ", round(frequency_ci_lower_non_cognate, 2), "]"
    ),
  ) %>% 
  relocate(
    language, version,
    n_identical_cognate,
    frequency_mean_identical_cognate,
    frequency_se_identical_cognate,
    frequency_ci_identical_cognate,
    matches("_non_identical"),
    matches("_non_cognate")
  ) %>% 
  mutate(language = ifelse(language=="Catalan", "CAT", "SPA")) %>% 
  select(-matches("sd|upper|lower"))

t <- d %>% 
  gt(groupname_col = "language") %>% 
  fmt_number(matches("mean|se")) %>% 
  tab_spanner("Identical cognates", columns = ends_with("_identical_cognate")) %>% 
  tab_spanner("Non-identical cognates", columns = ends_with("_non_identical_cognate")) %>% 
  tab_spanner("Non-cognates", columns = ends_with("_non_cognate")) %>% 
  cols_label(
    version = "List",
    n_identical_cognate = md("N"),
    n_non_identical_cognate = md("N"),
    n_non_cognate = md("N"),
    frequency_mean_identical_cognate = md("M"),
    frequency_mean_non_identical_cognate = md("M"),
    frequency_mean_non_cognate = md("M"),
    frequency_se_identical_cognate = md("SD"),
    frequency_se_non_identical_cognate = md("SD"),
    frequency_se_non_cognate = md("SD"),
    frequency_ci_identical_cognate = md("95% *CI"),
    frequency_ci_non_identical_cognate = md("95% CI"),
    frequency_ci_non_cognate = md("95% CI")
  ) %>% 
  cols_width(
    starts_with("frequency_ci_") ~ px(100),
    TRUE ~ px(40)
  ) %>% 
  tab_style(
    cell_text(style = "italic"),
    cells_column_labels(everything())
  ) %>% 
  tab_style(
    cell_text(weight = "bold"),
    cells_column_spanners(everything())
  )

gtsave(t, here("Figures/item_frequency.png"))

@@

```{r n_responses, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
n_obs <- nrow(responses)
sum_obs <- responses %>%
  group_by(te) %>%
  summarise(n = n(), .groups = "drop") %>%
  summarise(
    mean = mean(n),
    sd = sd(n),
    min = min(n),
    max = max(n)
  )

We gathered `r printnum(n_obs, big.mark = ",")` observations, with each observation corresponding to a single reponse of one participant to a single item. Translation equivalents (TEs) received an average of `r printnum(sum_obs$mean, digits = 2)` responses (*SD* = `r printnum(sum_obs$sd, digits = 2)`, *Min* = `r printnum(sum_obs$min, digits = 2)`, Max = `r printnum(sum_obs$max, digits = 2)`), both languages summed together. To investigate our hypotheses, we modelled the probability of children understanding or producing TEs using a logistic regression model with a Bernoulli distribution and a logit link function [@agresti2003categorical]. We included several predictors of interest as fixed effects to adjust this probability to participants' and items' characteristics, namely:

* *Age* of participant in months (`age_center`), calculated as the difference in days between participants' birth date and questionnaire completion divided by 30, chunked into bins of two months width, and centred around the mean.
* *Lexical frequency* of item (`frequency_center`), Relative lexical frequencies (frequency per million words, $fpmw$) were extracted from the CHILDES data base, using the `childesr` R package [@braginsky2018childesr], and were transformed to Zipf scores [@van2014subtlex; @zipf1949human] ($log10(fpmw)+3$). Missing frequencies were imputed by multiple imputation using the `mice` R package [van2011mice]. The resulting scores were then centred around the mean.
* *Degree of exposure* to the language (`doe_center`), computed as the proportion of the time the child is exposed to the language the item belongs to. For instance, for a response to the item `perro` (Spanish), the degree of exposure would correspond to the participant's relative exposure to Spanish (ranging from 0 to 1). Before entering the analyses, this variable was centred it around the mean and divided by 10, so that the values of the regression coefficients of this predictor correspond to a 10% increase in exposure to the language of the item.
* *Cognateness*, a categorical variable indicating whether the Catalan and Spanish forms of the TE are identical cognates (IC), non-identical cognates (NIC), or non-cognates (NC), as rated by a trained Spanish-Catalan bilingual linguist. We sum-coded two contrasts [@schad2020capitalize]: `cognate1`, comparing IC and NIC (both coded as +0.25) with NC (coded as -0.5) and `cognate2`, comparing IC (coded as +0.5) with NIC (coded as -0.5).

We first fit a base model that only included the main effects of `age_center` and `frequency_center`  (Model 1) [see @braginsky2019consistency; @jones2019children; @kachergis2021toward for similar approaches]. We then entered our predictors of interest by adding `doe_center` (Model 2) and then `cognate` and its interaction with `doe_center` (Model 3). To account for the hierarchical structure of the data (some responses to different TEs corresponded to the same participant, and some responses from different participants corresponded to the same TE )every model included *Participants* (`id`) and *TEs* (`te`) as grouping variables (i.e. random effects), with all possible random intercepts and slopes where appropriate [@barr2013random] (see Appendix for more details of the model).

We adopted a Bayesian approach toward statistical inference. This approach allows to (1) incorporate previous domain knowledge into the inference process implementing the Bayes theorem, and (2) to quantify the uncertainty associated to the estimated parameters in our model [@mcelreath2018statistical]. 
We fit the models using the R package `brms` [@burkner2017brms], an R interface to the probabilistic language Stan [@carpenter2017stan]. We run four sampling chains with 1,000 iterations each, including 500 warm-up iterations per chain. All R-hat values were smaller than 1.1, indicating that the chains converged successfully in all cases [@gelman1992inference] (see Appendix for an extended description of model diagnostics).

We then compared comprehension and production models separately using leave-one-out cross-validation (LOO-CV) [@vehtari2017practical]^[Due to the high computational cost associated with our large dataset, we used a sub-sampling approach for performing Bayesian LOO with 1,000 samples [@magnusson2019bayesian]]. This method computes, for a given model, the sum of the log scores of the posterior predictive distributions that result from removing one data-point at a time, providing an estimate of the model fit. The LOO estimates are adjusted for the number of parameters estimated in the model, therefore accounting for overfitting. We compared the expected log posterior density (ELPD) of each each model to test which one fitted the data the best.

@@

```{r coefs, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

coef_comp <- as.data.frame(fixef(fits_comp$fit_2))
coef_prod <- as.data.frame(fixef(fits_prod$fit_2))

For both comprehension and production, the extended model--which included all main effects and interactions--fitted data the best: it showed the smallest absolute expected log predictive density. This estimate is several times larger than its standard error indicating high predictive accuracy. Table 3 summarises the LOO scores of the fitted models. In the next sections we report the fixed effects of the comprehension and production models separately. We report the mean of each coefficient's posterior distribution along with standard errors and 95% credible intervals (CrI) in Tables 4 and 5. For interpretability, we report the derivative of the logistic function for each coefficient^[The derivative is calculated as $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$. This value corresponds to the slope of the logistic curve at the midpoint, where it is steepest.] for each coefficient, which indicates the maximum difference in probability corresponding to one unit different in the input variable [@gelman2020regression].

```{r loo_prod, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Leave-one-out cross-validation. The expected log predictive density (*ELPD*) indicates the sum of the likelihood of each observation, given the model when fitted without such obervation. Values close to 0 indicate that the model fits data better. Standard errors (*SE*) indicate the uncertainty around the computed ELPD in each model. A small *SE* (in realtion to the magnitde of the *ELPD* indicates less uncertainty around the ELPD estimate."}

l_comp <- loo_comp %>% 
  as_tibble() %>% 
  select(-matches("p_loo|looic")) %>% 
  mutate_all(as.numeric) %>% 
  mutate(
    type = "Comprehension",
    model = paste0("Model ", 2:0)
  ) %>% 
  relocate(model, type, matches("elpd_loo"), matches("diff"))

>>>>

bibliography      : ["references.bib"]

@@

knitr::opts_chunk$set(
    echo = FALSE,
    fig.align = "center",
    message = FALSE,
    warning = FALSE,
    cache.extra = knitr::rand_seed,
    out.width = "100%",
    dpi = 800,
    results = "asis"
)
```

@@

# load packages
library(bayesplot)
library(brms) 
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(quarto)
library(rmarkdown)
library(scales) 
library(stringdist)
library(stringr)
library(tidybayes) 
library(tidyr)

# load and process objects
tar_load_everything()
participants <- participants %>% 
    left_join(select(multilex_data$logs, id, time, edu_parent))
items <- items %>% 
    left_join(select(pool, item, language, category, class))

# set global options
options(
    knitr.kable.NA = "-",
    knitr.duplicate.label = "allow",
    ggplot2.discrete.fill = ggsci::pal_d3()(4),
    ggplot2.discrete.colour = ggsci::pal_d3()(4),
    ggplot2.continuous.fill = ggplot2::scale_color_gradient,
    ggplot2.continuous.colour = ggplot2::scale_color_gradient
)

# set custom ggplot2 theme
theme_set(theme_custom())

# resolve namespace conflicts
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
```

@@

```{r participants, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
n_participants <- length(unique(participants$id))
dates <- as_date(range(participants$time_stamp, na.rm = TRUE))
```

@@

```{r participants-lp, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Participant sample size by age and degree of exposure to a second language. Information about DoE was provided by families when filling the questionnaire. For illustration purposes this table, DoEs were binned into 10\\% wide bins, and ages (in months) were round to the nearest integer. Hyphens indicate that no participants from that specific combination of age filled the questionnaire."}
participants %>% 
    mutate_at(
        vars(starts_with("doe_")), 
        ~floor(.*10)/10
    ) %>% 
    mutate(age = cut_width(floor(age), 4)) %>% 
    count(age, doe_catalan) %>% 
    arrange(-doe_catalan) %>% 
    pivot_wider(names_from = age, values_from = n) %>% 
    gt() %>% 
    fmt_missing(everything(), everything(), missing_text = "--") %>% 
    cols_label(
        doe_catalan = "DoE Catalan"
    ) %>%
    fmt_percent("doe_catalan", decimals = 0) %>% 
    tab_spanner("Age (months)", 2:8)
```

We use the highest educational attainment of parents or caretakers as a proxy of participants' socio-economic status (SES), which families self-reported in the questionnaire by filling two items asking for the educational attainment of each parent or caretaker, with the following available options: *No education*, *Primary*, *Secondary*, *Complementary*, *Vocational*, and *University*, in line with the current educational system in Spain. Most families reported university studies (`r table(participants$edu_parent)["University"]`, `r table(participants$edu_parent)["University"]/length(unique(participants$id))`), followed by families were the highest educational attainment were vocational studies (`r table(participants$edu_parent)["Vocational"]`, `r percent(table(participants$edu_parent)["Vocational"]/length(unique(participants$id)))`), complementary studies (`r table(participants$edu_parent)["Complementary"]`,  `r percent(table(participants$edu_parent)["Complementary"]/length(unique(participants$id)))`), secondary education (`r table(participants$edu_parent)["Secondary"]`, `r percent(table(participants$edu_parent)["Secondary"]/length(unique(participants$id)))`, `r percent(table(participants$edu_parent)["Primary"]/length(unique(participants$id)))`), no formal education (`r table(participants$edu_parent)["No education"]`, `r percent(table(participants$edu_parent)["No education"]/length(unique(participants$id)))` and primary education (`r table(participants$edu_parent)["Primary"]`, <1%). Highest parental educational attainment was similar across participants with different DoE to a second language (see Appendix).

@@

```{r items, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
n_categories <- nrow(distinct(multilex_data$pool, category))
n_te <- length(unique(multilex_data$pool$te))
n_items <- length(unique(pool$item))
n_item_language <- count(pool, language) %>%
    pull(n) %>% 
    set_names(c("catalan", "spanish"))
```

The questionnaire was implemented on-line using the formR platform [@arslan2020formr], and was structured in three blocks: a (1) language questionnaire, a (2) demographic survey, and a (3) Catalan and a Spanish vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. The Catalan inventory contained `r n_item_language["catalan"]` items and the Spanish inventory contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., whenever *gos* [dog] was included in the Catalan inventory, the word *perro* was included in the Spanish inventory), roughly following a one-to-one mapping. In case two translation equivalents were possible for a given word, both were inluded as separate items (e.g., Catalan *acabar* [*to finish*] and Spanish *acabar* and *terminar*), or merged them into a single item (e.g., Spanish *mono* [*monkey*] and Catalan *mono/mico*). We included items from a diverse sample of `r n_categories` semantic/functional categories (see Appendix 1). For the analyses included in this study, we excluded items from the adverbs, auxiliary words, connectives, interjections and games and routines categories, so that only data from content words (nouns, adjectives, and verbs) were used.

For each word in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Some families filled a long version of the vocabulary checklists (`r format(n_te, big.mark = ",")` translation equivalents; `r format(n_item_language["catalan"], big.mark = ",")` items in Catalan, `r format(n_item_language["spanish"], big.mark = ",")` items in Spanish), while others filled a shorter version (~400 translation equivalents, ~400 items in Catalan, ~400 items in Spanish). These last families were randomly allocated into one of four different subsets of the complete list of items. These lists were designed so that each contained a representative sub-sample of the items from the complete list. Semantic/functional categories with less than 16 items--thus resulting in less than four items after dividing it in four lists--were not divided in the short version of the questionnaire: all of their items were included in the four lists. Another subset of items that were part of the trial lists of some experiments in the lab were also included in all versions. Table 2 in Appendix 1 shows the distribution of items across questionnaire versions.

To analyse the data, we considered responses to items corresponding to nouns, verbs, and adjectives [@fourtassi2020]. We excluded multi-word items (e.g., *barrita de cereales* [cereal bar]), items that included more than one word-form (e.g., *mono / mico*), and words with missing lexical frequency scores.

```{r item-summary, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=7, fig.cap="Lexical frequencies. Mean, standard error, and 95\\% confidence interval of lexical frequencies of items included in the Catalan and Spanish lists, reported separately for identical cognates, non-identical cognates, and non-cognates."}
items %>%
    unnest(list) %>% 
    group_by(language, list) %>% 
    summarise_at(
        vars(freq, n_phon, lv), 
        lst(n = ~sum(!is.na(.)), mean, sd, min, max)
    ) %>% 
    select(-c(n_phon_n, lv_n)) %>% 
    rename(n = freq_n) %>% 
    gt(groupname_col = "language") %>% 
    fmt_number(matches("freq_|n_phon_mean|n_phon_sd")) %>% 
    fmt_percent(matches("lv_")) %>% 
    tab_spanner("Frequency (Zipf)", columns = starts_with("freq_")) %>% 
    tab_spanner("# Phonemes", columns = starts_with("n_phon_")) %>% 
    tab_spanner("Levenshtein", columns = starts_with("lv_")) %>% 
    cols_merge_range(col_begin = "freq_min", col_end = "freq_max") %>% 
    cols_merge_range(col_begin = "n_phon_min", col_end = "n_phon_max") %>% 
    cols_merge_range(col_begin = "lv_min", col_end = "lv_max") %>% 
    cols_label(
        list = "",
        n = md("*N*"),
        freq_mean = md("Mean"),
        freq_sd = md("*SD*"),
        freq_min = md("Range"),
        n_phon_mean = md("Mean"),
        n_phon_sd = md("*SD*"),
        n_phon_min = md("Range"),
        lv_mean = md("Mean"),
        lv_sd = md("*SD*"),
        lv_min = md("Range")
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(everything())
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(everything())
    )
```

@@

```{r n_responses, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
n_obs <- nrow(df)
sum_obs <- df %>%
    group_by(te) %>%
    summarise(n = n(), .groups = "drop") %>%
    summarise(
        mean = mean(n),
        median = median(n),
        sd = sd(n),
        min = min(n),
        max = max(n)
    )
```

We gathered `r printnum(n_obs, big.mark = ",")` observations, with each observation corresponding to a single response of one participant to a single item. Translation equivalents (TEs) received a median of `r printnum(sum_obs$median, digits = 2)` responses (*Min* = `r printnum(sum_obs$min, digits = 2)`, *Max* = `r printnum(sum_obs$max, digits = 2)`), both languages summed together.

We modelled the probability of answering each response category (*No* < *Understands* < *Understands and Says*) using a mixed effects model. This allowed us to simultaneously estimate the impact of item- and participant-level predictors on the probability of word comprehension and production. We included the following predictors: age of the participant in months, calculated as the difference in months between participants' birth date and questionnaire completion ($Age$), word lexical frequency, extracted from the CHILDES database using the `childesr` R package [[@braginsky2018childesr]] and expressed in Zipf scores, ($log10(fpmw)+3$) [@van2014subtlex; @zipf1949human] ($Frequency$), number of phonemes included in the IPA phonological transcription of the word-form ($Phonemes$), participant's degree of exposure to the language the item belongs to ($DoE$)^[For instance, for a response to the item `perro` (Spanish), the degree of exposure would correspond to the participant's relative exposure to Spanish (ranging from 0 to 1).], and the Levenshtein distance between the IPA transcriptions of the translation equivalent in Catalan and in Spanish, as a measure of their phonological similarity ($Levenshtein$) [@floccia2018methods].

We included $Age$, $Frequency$, $Phonemes$, $DoE$, and $Levenshtein$ as fixed effects. We also included a three-way interaction between $Age$, $DoE$, and $Levenshtein$ and its corresponding two-way interactions. We added participants and items as crossed random effects including all intercepts and slopes when appropriate [@barr2013maximal]. This model structure, including word- and participant-level properties to estimate vocabulary acquisition trajectories was discussed by @kachergis2021toward in the context of explanatory item response models. Our model differs from their proposal in that  1) while they modelled global trajectories exclusively, we modelled individual trajectories as well, 2) our outcome variable is ordinal in contrast to the binary outcome used in their study, 3) we propose a Bayesian approach. Our motivation for a Bayesian approach towards statistical inference is that (1) it allows us to incorporate previous domain knowledge to improve estimation accuracy, and (2) to quantify the uncertainty associated with the estimated parameters in our model [@mcelreath2018statistical].

The Bayesian approach implies using Bayes’ rule to assign a probability to possible parameter values (i.e., posterior distribution), based on previous knowledge about such probability distribution (prior distribution) and our data (likelihood). We used Markov chain Monte Carlo (MCMC) to approximate the posterior distribution via sampling. We run `r dim(model_fit_7$fit)[2]` sampling chains with `r dim(model_fit_7$fit)[1]` iterations each, including `r dim(model_fit_7$fit)[1]/2` warm-up iterations per chain. All $\hat{R}$ values were smaller than 1.1, indicating that the chains converged successfully in all cases [@gelman1992inference] (see Appendix for an extended description of model structure and diagnostics).

We compared a base model–only including fixed and random intercepts–to increasingly more complex models, introducing predictors in the following order: $Age$, $Frequency$, $Phonemes$, $DoE$, $Levenshtein$, two-, and three-way interactions [see @braginsky2019consistency; @jones2019children; @kachergis2021toward for similar approaches]. We compared the predictive validity of the models using leave-one-out cross-validation (LOO-CV), and selected the model with the smallest expected log posterior density (ELPD) [@vehtari2017practical]^[Due to the high computational cost associated with our large dataset, we used a sub-sampling approach for performing Bayesian LOO with 1,000 samples [@magnusson2019bayesian]]. Finally, we explored the posterior distribution of each parameter in the selected model. We report the mean and 95% credible interval of the posterior distribution of the parameters of interest. We performed follow-up tests on interactions whose 95% credible interval excluded 0 by comparing the 95% CrI of their expected posterior marginal means using the `tidybayes` R package [@R-tidybayes]. Data processing and visualisation was done in R [@R-base] using the `tidyverse` family of packages [@R-tidyverse].

@@

```{r coefs, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
coefs <- as.data.frame(fixef(model_fit_7)) %>% 
    clean_names()

draws_fixed <- gather_draws(model_fit_7, `b_.*`, regex = TRUE)
```

Table 2 shows the summary of the posterior distribution of the fixed coefficients of the model.

```{r results-fixed, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
str_repl <- c(
    "b_age_std:doe_std:lv_std" = "Age \u00d7 DoE \u00d7 Levenshtein",
    "b_age_std:doe_std" = "Age \u00d7 DoE",
    "b_age_std:lv_std" = "Age \u00d7 Levenshtein",
    "b_doe_std:lv_std" = "DoE \u00d7 Levenshtein",
    "b_lv_std" = "Levenshtein (+1 SD)",
    "b_doe_std" = "DoE (+10%)",
    "b_n_phon_std" = "# Phonemes (+1 SD)",
    "b_freq_std" = "Frequency (+1 SD)",
    "b_age_std" = "Age (+1 month)",
    "b_Intercept[1]" = "Intercept (Production)",
    "b_Intercept[2]" = "Intercept (Comprehension)"
)

draws_fixed %>%
    mutate(
        .value = ifelse(
            grepl("Intercept", .variable),
            inv_logit_scaled(.value), 
            .value/4
        ),
        .variable = factor(.variable, levels = names(str_repl), labels = str_repl)
    ) %>% 
    mean_qi(.width = c(0.95, 0.89)) %>% 
    pivot_wider(names_from = .width, values_from = c(.value, .lower, .upper)) %>% 
    select(-c(.interval, .point, .value_0.89)) %>% 
    arrange(desc(.variable)) %>% 
    gt() %>% 
    fmt_percent(2:6) %>% 
    cols_merge(vars(".lower_0.95", ".upper_0.95"), pattern = "[{1}, {2}]") %>% 
    cols_merge(vars(".lower_0.89", ".upper_0.89"), pattern = "[{1}, {2}]") %>% 
    cols_label(
        .variable = md("Predictor"),
        .value_0.95 = md("Mean"),
        .lower_0.95 = md("95\\% *CrI*"),
        .lower_0.89 = md("89\\% *CrI*")
    ) %>% 
    tab_footnote(
        footnote = "Transformed using the inverse logit to get the average probability of correct response",
        locations = cells_body(
            columns = ".variable", 
            rows = grepl("Intercept", .variable)
        ) 
    ) %>% 
    tab_footnote(
        footnote = "Transformed using the divide-by-four rule to get the maximum change in probability of correct response, associated with a unit increase in this variable.",
        locations = cells_body(
            columns = ".variable",
            rows = !grepl("Intercept", .variable)
        )
    )  %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = everything())
    ) %>% 
    tab_header(title = "Model fixed coefficients", subtitle = "Posterior distribution") %>% 
    tab_style(
        cell_text(align = "left"),
        list(cells_body(), cells_column_labels(columns = everything()))
    )
```

The extended model--which included all main effects and interactions--fitted data the best: it showed the smallest absolute expected log predictive density. This estimate is several times larger than its standard error indicating high predictive accuracy. Table 3 summarises the LOO scores of the fitted models. In the next sections we report the fixed effects of the comprehension and production models separately. We report the mean of each coefficient's posterior distribution along with standard errors and 95% credible intervals (CrI) in Tables 4 and 5. For interpretability, we report the derivative of the logistic function for each coefficient^[The derivative is calculated as $\hat{\beta_j}/4$, where $\hat{\beta_j}$ is the estimated mean of the posterior distribution of coefficient $j$. This value corresponds to the slope of the logistic curve at the midpoint, where it is steepest.] for each coefficient, which indicates the maximum difference in probability corresponding to one unit different in the input variable [@gelman2020regression].
